### 2017-03-14 term_paper/paper.tex
Start of term paper.

\documentclass[12pt]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{setspace}

\title{Term Paper}
\author{Philip Westrich}
\date{April 25, 2017}

\begin{document}

\maketitle

\vspace{-0.3in}
\noindent
\rule{\linewidth}{0.4pt}
\doublespacing

This is my paper. It isn't very long.

\end{document}

### 2017-03-24 term_paper/paper.tex
Got the start of the paper going. It should at least be in the right format.
Abstract exists now. So does bibliography.

@@
-\documentclass[12pt]{article}
-\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
-\usepackage{setspace}
+\documentclass[format=acmsmall, review=false, screen=true]{acmart}
 
-\title{Term Paper}
-\author{Philip Westrich}
-\date{April 25, 2017}
+\usepackage{booktabs} % For formal tables
 
+\usepackage[ruled]{algorithm2e} % For algorithms
+\renewcommand{\algorithmcfname}{ALGORITHM}
+\SetAlFnt{\small}
+\SetAlCapFnt{\small}
+\SetAlCapNameFnt{\small}
+\SetAlCapHSkip{0pt}
+\IncMargin{-\parindent}
+
+% Copyright
+\setcopyright{acmlicensed}
+
+% Document starts
 \begin{document}
+% Title portion. Note the short title for running heads 
+\title[Short Title]{Effects of Solid State Drives on Database Performance}  
+\author{Philip M. Westrich}
+\affiliation{
+  \institution{Tennessee Technological University}
+  \streetaddress{1 William L. Jones Drive}
+  \city{Cookeville} \state{TN} \postcode{38501} \country{USA}
+}
+
+\begin{abstract}
+
+    In the last decade or so, a new form of flash-based persistent storage, known as solid state drives (SSDs), have begun 
+    to replace traditional mechanical hard disk drives (HDDs). They are orders of magnitude faster, lighter, more energy 
+    efficient, and less prone to sudden impact than HDDs. As their prices fall and storage capacities increase, they have 
+    taken more of a hold in both the consumer and enterprise market. These new SSDs behave differently than HDDs, which can 
+    have a significant effect on any read or write heavy workload, such as database management systems. In this paper, we explore 
+    what these effects are and their proposed solutions.
+
+\end{abstract}
 
@@
 
-\vspace{-0.3in}
-\noindent
-\rule{\linewidth}{0.4pt}
-\doublespacing
+\section{Introduction}
 
-This is my paper. It isn't very long.
+Text goes here.
+
+\section{Overview of SSDs}
+
+Text goes here.
+
+\section{Effects on database systems}
+
+Text goes here.
+
+\section{Solutions to these effects}
+
+Text goes here.
+
+\section{Conclusions}
+
+Text goes here.
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{bibliography}
 

### 2017-04-09 term_paper/paper.tex
Added readme. Wrote introduction for term paper. One section down, four to go.

@@
 
-Text goes here.
+For years, the gap in performance between processors and persistent storage has continually widened. Disk latency for 
+traditional hard disk drives (HDDs) has improved by approximately ten percent per year, while we are just now beginning 
+to see the end of Moore's Law for processors. \cite{Xie2011}
+
+Traditional HDDs rely on mechanical parts, and therefore are destined to fail eventually, take relative eons to read 
+or write data, and are prone to multiuple methods of failure, such as sudden force, vibration, or simply mechanical 
+failure. They also are power inefficent and generate pleny of heat. \cite{Xie2011}
+
+With the introduction of solid state drives (SSDs), we can begin to address many of the aforementioned issues, as they 
+are not susceptible to the same problems. SSDs are lighter, faster, more energy efficient, and resistant to blunt force.
+\cite{Xie2011}
+
+However, they are not completely free from issue. At the time of writing, SSDs are still much more expensive per gigabyte 
+than their HDD counterparts, though they promise to reach equivalence within a few years. They also have much different 
+performance characteristics than HDDs, most notably, usually being worse at small random writes. They also tand to become 
+slower as they fill up, and have a limited number of read/write/erase cycles they can endure before failing. 
+\cite{Xie2011, Dumitru2007}
+
+Overall, SSDs are better than their predecessor, and will most likely come to replace them in the coming years. Therefore, 
+applications that are heavily reliant on their storage medium, such as databases, must come prepared for it. In this paper, 
+we will explore what effects these performance differences have as well as the solutions that have been proposed.
 

### 2017-04-10 term_paper/paper.tex
Added a couple more papers. Wrote a couple more paragraphs.

@@
 than their HDD counterparts, though they promise to reach equivalence within a few years. They also have much different 
-performance characteristics than HDDs, most notably, usually being worse at small random writes. They also tand to become 
+performance characteristics than HDDs, most notably, usually being worse at small random writes. They also tend to become 
 slower as they fill up, and have a limited number of read/write/erase cycles they can endure before failing. 
@@
 
-\section{Overview of SSDs}
+\section{Overview of solid state drives}
 
-Text goes here.
+Though solid state drives are designed to emulate the behavior of previous block devices, under the hood, they operate 
+quite differently. \cite{Lee2008, Cornwell2012, Micheloni2013, MatejFucek2014} In this section, we explain the general 
+architecture of a solid state drive and explain these differences.
+
+Solid state drives consist of three major components: an array of non-volatile flash memory, a host interface, and a 
+microcontroller that bridges the gap between the two.
+
+\subsection{NAND flash}
+
+Most SSDs today use a type of electrically erasable programmable read-only memory (EEPROM) known as NAND flash memory. 
+NAND is an abbreviation for 'not-and', a type of logic gate. These NAND cells are arranged into a grid pattern. These 
+cells can only be accessed in certain patterns dependent on the particular drive's arrangement of those cells. 
+\cite{Cornwell2012, Micheloni2013}
+
+Several of these grids, usually one to four, are placed together into what is known as a die. Each die can read somewhere 
+around 400 MB/sec, while they can only write at around 20MB/sec. This is due to the much more complicated write method 
+they employ. However, they have much lower latencies for these operations, typically measured in microseconds, and in 
+order to meet capacity and speed requirements, multiple dies are packaged inside one SSD. \cite{Cornwell2012, Micheloni2013}
+
+SSDs use a write-verify method to program, or write to, the cells. In this procedure, the cells are first erased, then 
+a high voltage signal is applied to them until they reach the appropriate value. As the cells hold more bits, this 
+process gets more difficult and slows down, as the voltages applied to the cells must become more fine-grained and precise.
+\cite{Cornwell2012, Micheloni2013}
+
+\subsection{SSD controller}
+
+The controller's job is twofold. First, it must translate the operations requested by the host machine into operations 
+that be performed on the NAND flash. Second, it must do this in a manner such that wear on the device is minimized and 
+even, and the time needed to perform the requested operations is minimized.
+
+To carry out these tasks, SSDs have a small computer, known as a microcontroller. The microcontroller is also paired with 
+specialized hardware to speed up its computations.
 

### 2017-04-11 term_paper/paper.tex
Lots more paragraphs.

@@
 % Title portion. Note the short title for running heads 
-\title[Short Title]{Effects of Solid State Drives on Database Performance}  
+\title[Effects of SSDs on Databases]{Effects of Solid State Drives on Database Performance}  
 \author{Philip M. Westrich}
@@
 around 400 MB/sec, while they can only write at around 20MB/sec. This is due to the much more complicated write method 
-they employ. However, they have much lower latencies for these operations, typically measured in microseconds, and in 
-order to meet capacity and speed requirements, multiple dies are packaged inside one SSD. \cite{Cornwell2012, Micheloni2013}
+they employ. However, they have a much lower latency for these operations, typically measured in microseconds, and in 
+order to meet capacity and speed requirements, multiple dies are packaged inside one SSD. The controller will often 
+read and write to multiple dies at once in order to reach performance goals. \cite{Cornwell2012, Micheloni2013}
 
@@
 
-The controller's job is twofold. First, it must translate the operations requested by the host machine into operations 
-that be performed on the NAND flash. Second, it must do this in a manner such that wear on the device is minimized and 
-even, and the time needed to perform the requested operations is minimized.
+The controller has many jobs. First, it must present to the host device the same interface HDDs do. Second, it must 
+translate the commands issued to it to operations on the NAND flash. Third, it must wear the NAND in the drive evenly 
+to prevent premature failure. Fourth, it must deal with the higher error rate that the NAND flash has over HDDs. 
+\cite{Cornwell2012, Micheloni2013, MatejFucek2014}
 
-To carry out these tasks, SSDs have a small computer, known as a microcontroller. The microcontroller is also paired with 
-specialized hardware to speed up its computations.
+To carry out these tasks, SSDs have a small computer, known as a microcontroller, embedded within them. The microcontroller 
+is also paired with specialized hardware to speed up some of its computations, as well as some amount of DRAM to cache and 
+buffer events. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+
+The controller must present to the host device the same interface other block devices do. However, the NAND flash is not 
+accessed in the same way; instead, it has its own filesystem it maintains, known as the flash translation layer, or FTL. 
+The FTL is used to perform the translation from the logical blocks the host requests to the pages and blocks within the 
+NAND flash. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+
+Because cells must be erased before they can be written to, new data is always written to known empty pages, and the old 
+pages are marked as garbage in the FTL. Therefore, the actual locations of data may move across the drive as it is written 
+to, unlike HDDs. The FTL must maintain a list of empty, written, and garbage pages, and keep track of the mapping from 
+the logical blocks requested by the host and their actual location on the SSD. The FTL also maintains the number of writes 
+a block has experienced as well as a list of bad blocks that have either started out defective from the factory, or have 
+worn out over time. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+
+When the controller is idle for some period of time, it will start performing garbage collection on the now invalid pages 
+in order to free up more space. Some SSDs, especially earlier ones, would degrade in performance as the drive filled up 
+and it ran out of free pages. The drive then had to immediately garbage collect to free up more, causing a delay in writes. 
+Modern SSDs will understate their capacity to the host system, that way there are always free pages on the disk. 
+\cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+
+NAND flash has a much higher error rate than the magnetic platters of HDDs. Therefore, all SSDs have some amount of 
+hardware-assisted error correction. Higher quality drives have more dedicated hardware, while cheaper ones may share 
+hardware for a larger number of pages. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}
 
@@
 
-Text goes here.
+For years, systems have assumed that HDDs are the form of persistent storage used, and are optimized for it. These 
+optimizations often are detrimental to SSD performance. In this section, we explain these assumptions and why they 
+may lead to suboptimal performance.
+
+\subsection{Physical differences}
+
+HDDs have multiple moving parts: the platters that spin at some speed, and the heads that move across them that read the 
+data from the platters. It takes time to move these parts through space. Often, to conserve power and lifespan, the drive 
+stops rotating its platters. When at a complete stop, it can take between four to ten seconds to bring the platters back 
+up to speed. This may not be an issue in consumer-grade applications, but it can be a matter of life and death in critical
+real-time systems. \cite{MatejFucek2014} The head of the drive must also move to the correct location to read from the 
+requested sector. \cite{Cornwell2012}
+
+Solid state drives have zero moving parts, hence their name. They do not require any spin up time, and their seek time is 
+both uniform and much, much smaller. An additional benefit is that they consume much less energy. 
+\cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+
+\subsection{Differences in access patterns}
+
+Current systems tend to assume that they will be placing their data on to HDDs, and tend to optimize for that; they try 
+to make up for the shortcomings of the physical media. They assume the location of the sectors requested are always in 
+the same location, and that lower addresses are towards the faster, outside of the drive, and higher addresses are towards 
+the inside. \cite{Cornwell2012}
+
+Because of the seek time added by moving the head across the drive and the assumption of block locations, filesystem drivers 
+will perform complicated math to determine how it can move the head the minimum distance for a sequence of reads and writes. 
+They will also perform \textit{I/O coalescing,} or merging many smaller requests into one larger one, since HDDs are much 
+better at large sequential operations. \cite{Cornwell2012}
+
+In order to minimize the number of requests made to the drive, filesystem drivers will typically not inform the drive that 
+blocks are freed; they simply mark in their own data structures that the blocks are not needed, and the data remains on 
+the disk until something comes to overwrite it. \cite{Cornwell2012}
+
+SSDs break many of these assumptions. Because the FTL takes the liberty of shuffling data around on the disk as it sees 
+fit, I/O coalescing and other location assumptions are meaningless, and the time and effort put in to them are wasted. 
+Not informing the drive of freed blocks harms the performance of SSDs, as they rely on having as many free pages as 
+possible. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+
+\subsection{What this means for databases}
+
+With the direction that computing is headed, data is invaluable. The databases in which this data is stored are completely 
+dependent on the storage medium in which they are backed.
 

### 2017-04-11 term_paper/paper.tex
More citations, more paragraphs. I think the end is in sight…

@@
 slower as they fill up, and have a limited number of read/write/erase cycles they can endure before failing. 
-\cite{Xie2011, Dumitru2007}
+\cite{Dumitru2007, Dirik2009, Xie2011}
 
@@
 Though solid state drives are designed to emulate the behavior of previous block devices, under the hood, they operate 
-quite differently. \cite{Lee2008, Cornwell2012, Micheloni2013, MatejFucek2014} In this section, we explain the general 
-architecture of a solid state drive and explain these differences.
+quite differently. \cite{Lee2008, Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016} In this section, we 
+explain the general architecture of a solid state drive and explain these differences.
 
@@
 cells can only be accessed in certain patterns dependent on the particular drive's arrangement of those cells. 
-\cite{Cornwell2012, Micheloni2013}
+\cite{Dirik2009, Cornwell2012, Micheloni2013, Chen2016}
 
@@
 order to meet capacity and speed requirements, multiple dies are packaged inside one SSD. The controller will often 
-read and write to multiple dies at once in order to reach performance goals. \cite{Cornwell2012, Micheloni2013}
+read and write to multiple dies at once in order to reach performance goals. 
+\cite{Dirik2009, Cornwell2012, Micheloni2013, Chen2016}
 
@@
 process gets more difficult and slows down, as the voltages applied to the cells must become more fine-grained and precise.
-\cite{Cornwell2012, Micheloni2013}
+\cite{Dirik2009, Cornwell2012, Micheloni2013, Chen2016}
 
@@
 to prevent premature failure. Fourth, it must deal with the higher error rate that the NAND flash has over HDDs. 
-\cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+\cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
@@
 is also paired with specialized hardware to speed up some of its computations, as well as some amount of DRAM to cache and 
-buffer events. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+buffer events. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
@@
 The FTL is used to perform the translation from the logical blocks the host requests to the pages and blocks within the 
-NAND flash. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+NAND flash. \cite{Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
@@
 a block has experienced as well as a list of bad blocks that have either started out defective from the factory, or have 
-worn out over time. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+worn out over time. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
@@
 Modern SSDs will understate their capacity to the host system, that way there are always free pages on the disk. 
-\cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+\cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
@@
 hardware-assisted error correction. Higher quality drives have more dedicated hardware, while cheaper ones may share 
-hardware for a larger number of pages. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+hardware for a larger number of pages. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
@@
 
-With the direction that computing is headed, data is invaluable. The databases in which this data is stored are completely 
-dependent on the storage medium in which they are backed.
+With the direction that computing is headed, data is invaluable. Data-heavy applications, such as those that heavily use 
+online transaction processing (OLTP), require a data store that provides high performance and reliability.  The databases 
+in which this data is stored are completely dependent on the storage medium in which they are backed. Its characteristics 
+can seriously affect the end performance of the system. \cite{Lee2009, Xie2011, Micheloni2013, MatejFucek2014, Chen2016}
+
+OLTP systems often require many small, random writes, which are the one thing that SSDs are terrible at. A small write 
+requires rewriting the entire page with its changes to a separate location in the drive, then erasing the previous one, 
+which is a costly operation. Many of these mall writes performed close together may lead to the SSD running out of free 
+pages to write to, which would degrade performance tremendously. \cite{Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
+
+Other systems that have a more read-heavy workload will improve greatly with the introduction of SSDs, especially as the 
+randomness of those reads increases. Their only worry would then be the higher cost of large storage, which promises to 
+eventually fall to the price of current HDDs. \cite{Chen2016}
 
 \section{Solutions to these effects}
+
+Some have already explored replacing HDDs with SSDs in the enterprise space, and have proposed some solutions to the 
+problems that they introduce. In this section, we explain these solutions.
+
+\subsection{Hardware improvements}
+
+SSD manufacturers have made numerous advantage on their end to help mitigate what they are bad at. There are several 
+places apt for improvement, such as the FTL, or simply the amount of hardware included in the device. The more efficient 
+the FTL is at translating and ordering requests and the more hardware that allows for increased parallelism, the more 
+performance will come from the device. \cite{Xie2011, Chen2016}
+
+Many SSDs share hardware for both error correction and the write-verify cycle. Enterprise-grade devices tend to have 
+less hardware sharing, increasing the amount of parallelism; areas of the drive will not have to wait in line to use 
+the shared hardware. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
+
+An improved interface for the device will also increase SSD performance. Currently, SSDs use the same interface that 
+HDDs have employed, serial advanced technology attachment generation 3 (SATA-3), which is limited to somewhere around
+600 MB/sec of thourughput. While that may be sufficent for HDDs, SSDs can easily outpace that. \cite{Micheloni2013}
+
+Fortunately, there are faster interfaces present on modern day systems. The peripheral component interconnect express 
+(PCIe) interface is used in all modern-day systems to connect important devices together. In fact, the SATA controller 
+is often itself a PCIe device. PCIe promises transfer speeds anywhere from two to eight times that of SATA dependent on 
+the number of lanes dedicated to the device. It also supports duplex and device-to-device communication, eliminating 
+overhead. \cite{Micheloni2013}
+
+Moving SSDs to PCIe would allow for much higher throughput, lower latency, and simplification of the FTL; it would no 
+longer need to emulate a traditional block device. Other PCIe devices could also directly interact with the SSD to move 
+data, reducing communication overhead with the CPU. \cite{Micheloni2013}
+
+\subsection{Software improvements}
 

### 2017-04-12 term_paper/paper.tex
More references, more paragraphs.

@@
 
-Text goes here.
+There are also a number of software improvements that can be made to take advantage of what flash-based storage is good 
+at and helping to mitigate what they are bad at, much like what current filesystem drivers have done with magnetic media.
+
+Operating systems file drivers will need to become aware of flash-based media. We have seen this recently, for example, 
+with Apple's introduction of the Apple File System (APFS). It promises to bring with it a number improvements, most notably
+in the case of this papaer, being optimized to work with flash-based storage. \cite{AppleComputer2017}
+
+Systems that are data-heavy will need to attempt to optimize their access patterns for SSDs rather than HDDs.
 
@@
 
-Text goes here.
+Overall, solid state drives are a huge performance improvement over rotational magnetic media. They help greatly in closing 
+the performance gap between processors and storage that we have seen widen in the last several decades. It will only be a 
+matter of time before SSDs reach the same data density and cost per gigabyte of HDDs. \cite{Chen2016}
+
+However, SSDs introduce new problems in applications due to the vastly different behaviors that they exhibit. Both software 
+and hardware will need to start accommodating for these differences in order to reduce bottlenecks in data-intensive 
+applications, such as database management systems. We have already seen this with the works in \cite{AppleComputer2017}, 
+and more is sure to come.
 

### 2017-05-01 term_paper/paper.tex
Paper might be finished?

@@
     efficient, and less prone to sudden impact than HDDs. As their prices fall and storage capacities increase, they have 
-    taken more of a hold in both the consumer and enterprise market. These new SSDs behave differently than HDDs, which can 
-    have a significant effect on any read or write heavy workload, such as database management systems. In this paper, we explore 
-    what these effects are and their proposed solutions.
+    become much more popular in both the consumer and enterprise market. These new SSDs have different performance characteristics
+    than mechanical drives, which can impact the performance on storage-heavy applications, such as databases. In this paper, 
+    we explore what these impacts are and their proposed solutions.
 
@@
 For years, the gap in performance between processors and persistent storage has continually widened. Disk latency for 
-traditional hard disk drives (HDDs) has improved by approximately ten percent per year, while we are just now beginning 
-to see the end of Moore's Law for processors. \cite{Xie2011}
+traditional hard disk drives (HDDs) has improved by approximately ten percent per year, but processor speed has nearly 
+doubled each year according to Moore's Law until just recently. \cite{Xie2011}
 
-Traditional HDDs rely on mechanical parts, and therefore are destined to fail eventually, take relative eons to read 
-or write data, and are prone to multiuple methods of failure, such as sudden force, vibration, or simply mechanical 
-failure. They also are power inefficent and generate pleny of heat. \cite{Xie2011}
+Traditional HDDs rely on mechanical parts, which degrade over time, are prone to multiple methods of failure, such as 
+sudden force, vibration, or simply mechanical failure. They also are inefficient at their jobs; they read and write data 
+slowly while using lots of energy and generating lots of heat. \cite{Xie2011}
 
-With the introduction of solid state drives (SSDs), we can begin to address many of the aforementioned issues, as they 
-are not susceptible to the same problems. SSDs are lighter, faster, more energy efficient, and resistant to blunt force.
-\cite{Xie2011}
+With the introduction of solid state drives (SSDs), they can address many of the aforementioned issues. SSDs are lighter, 
+faster, more energy efficient, and resistant to blunt force. \cite{Xie2011}
 
-However, they are not completely free from issue. At the time of writing, SSDs are still much more expensive per gigabyte 
-than their HDD counterparts, though they promise to reach equivalence within a few years. They also have much different 
-performance characteristics than HDDs, most notably, usually being worse at small random writes. They also tend to become 
-slower as they fill up, and have a limited number of read/write/erase cycles they can endure before failing. 
+However, they are not completely free from issue. Currently, SSDs are still much more expensive per gigabyte than their 
+HDD counterparts, though they promise to reach equivalence within a few years. SSDs also have much different performance 
+characteristics than HDDs. Most notably, SSDs are usually slower at small random writes. SSDs also tend to become slower 
+as they fill up, and have a limited number of read/write/erase cycles they can endure before failing. 
 \cite{Dumitru2007, Dirik2009, Xie2011}
 
-Overall, SSDs are better than their predecessor, and will most likely come to replace them in the coming years. Therefore, 
-applications that are heavily reliant on their storage medium, such as databases, must come prepared for it. In this paper, 
-we will explore what effects these performance differences have as well as the solutions that have been proposed.
+Overall, SSDs are better than their predecessor, and will most likely replace HDDs in the coming years. The adoption of 
+SSDs will impact applications that are heavily reliant on their storage medium, such as databases. In this paper, we will 
+explore what effects these performance differences have as well as the solutions that have been proposed.
 
@@
 
-Though solid state drives are designed to emulate the behavior of previous block devices, under the hood, they operate 
-quite differently. \cite{Lee2008, Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016} In this section, we 
-explain the general architecture of a solid state drive and explain these differences.
+Solid state drives are designed to emulate the behavior of previous block devices. However, they operate quite differently. 
+\cite{Lee2008, Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016} In this section, we explain the general 
+architecture of a solid state drive and explain these differences.
 
@@
 translate the commands issued to it to operations on the NAND flash. Third, it must wear the NAND in the drive evenly 
-to prevent premature failure. Fourth, it must deal with the higher error rate that the NAND flash has over HDDs. 
-\cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
+to prevent premature failure, which is known as wear leveling. Fourth, it must deal with the higher error rate that 
+the NAND flash has over HDDs. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
 To carry out these tasks, SSDs have a small computer, known as a microcontroller, embedded within them. The microcontroller 
-is also paired with specialized hardware to speed up some of its computations, as well as some amount of DRAM to cache and 
-buffer events. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
+is also paired with specialized hardware to speed up some of its computations, as well as some DRAM to cache and buffer 
+events. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
 The controller must present to the host device the same interface other block devices do. However, the NAND flash is not 
-accessed in the same way; instead, it has its own filesystem it maintains, known as the flash translation layer, or FTL. 
+accessed in the same way; instead, it has its own filesystem it maintains, known as the flash translation layer (FTL). 
 The FTL is used to perform the translation from the logical blocks the host requests to the pages and blocks within the 
@@
 Because cells must be erased before they can be written to, new data is always written to known empty pages, and the old 
-pages are marked as garbage in the FTL. Therefore, the actual locations of data may move across the drive as it is written 
-to, unlike HDDs. The FTL must maintain a list of empty, written, and garbage pages, and keep track of the mapping from 
-the logical blocks requested by the host and their actual location on the SSD. The FTL also maintains the number of writes 
-a block has experienced as well as a list of bad blocks that have either started out defective from the factory, or have 
-worn out over time. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
+pages are marked as garbage in the FTL. The actual locations of data may move across the drive as it is written to, unlike 
+HDDs. The FTL must maintain a list of empty, written, and garbage pages, and keep track of the mapping from the logical 
+blocks requested by the host and their actual location on the SSD. The FTL also maintains the number of writes a block 
+has experienced as well as a list of bad blocks that have either started out defective from the factory, or have worn out 
+over time. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
-When the controller is idle for some period of time, it will start performing garbage collection on the now invalid pages 
-in order to free up more space. Some SSDs, especially earlier ones, would degrade in performance as the drive filled up 
-and it ran out of free pages. The drive then had to immediately garbage collect to free up more, causing a delay in writes. 
+When the controller is idle for some period of time, it will start performing garbage collection on the marked invalid 
+pages in order to free up more space. Some SSDs, especially earlier ones, would degrade in performance as the drive filled 
+up and it ran out of free pages. The drive then had to immediately garbage collect to free up more, causing a delay in writes. 
 Modern SSDs will understate their capacity to the host system, that way there are always free pages on the disk. 
@@
 
-NAND flash has a much higher error rate than the magnetic platters of HDDs. Therefore, all SSDs have some amount of 
-hardware-assisted error correction. Higher quality drives have more dedicated hardware, while cheaper ones may share 
-hardware for a larger number of pages. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
+NAND flash has a much higher error rate than the magnetic platters of HDDs. All SSDs have some amount of hardware-
+assisted error correction to make up for this. Higher quality drives have more dedicated hardware, while cheaper ones 
+may share hardware for a larger number of pages. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
@@
 
-Solid state drives have zero moving parts, hence their name. They do not require any spin up time, and their seek time is 
-both uniform and much, much smaller. An additional benefit is that they consume much less energy. 
-\cite{Cornwell2012, Micheloni2013, MatejFucek2014}
+Solid state drives have zero moving parts; that is where the name solid-state comes from. They do not require any spin up 
+time, and their seek time is both uniform and much, much smaller. An additional benefit is that they consume much less 
+energy. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}
 
@@
 the same location, and that lower addresses are towards the faster, outside of the drive, and higher addresses are towards 
-the inside. \cite{Cornwell2012}
+the slower inside. \cite{Cornwell2012}
 
 Because of the seek time added by moving the head across the drive and the assumption of block locations, filesystem drivers 
-will perform complicated math to determine how it can move the head the minimum distance for a sequence of reads and writes. 
-They will also perform \textit{I/O coalescing,} or merging many smaller requests into one larger one, since HDDs are much 
-better at large sequential operations. \cite{Cornwell2012}
+will compute complicated heuristics to determine how it can move the head the minimum distance for a sequence of reads 
+and writes. They will also perform \textit{I/O coalescing,} or merging many smaller requests into one larger one, since 
+HDDs are much better at large sequential operations. \cite{Cornwell2012}
 
@@
 online transaction processing (OLTP), require a data store that provides high performance and reliability.  The databases 
-in which this data is stored are completely dependent on the storage medium in which they are backed. Its characteristics 
-can seriously affect the end performance of the system. \cite{Lee2009, Xie2011, Micheloni2013, MatejFucek2014, Chen2016}
+in which this data is stored are completely dependent on the storage medium that backs them. Its characteristics can 
+greatly impact the end performance of the system. \cite{Lee2009, Xie2011, Micheloni2013, MatejFucek2014, Chen2016}
 
-OLTP systems often require many small, random writes, which are the one thing that SSDs are terrible at. A small write 
-requires rewriting the entire page with its changes to a separate location in the drive, then erasing the previous one, 
-which is a costly operation. Many of these mall writes performed close together may lead to the SSD running out of free 
-pages to write to, which would degrade performance tremendously. \cite{Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
+OLTP systems often require many small, random writes, which are the one thing that SSDs tend to perform worse at over 
+HDDs. A small write requires rewriting the entire page with its changes to a separate location in the drive, then erasing 
+the previous one, which is a costly operation. Many of these mall writes performed close together may lead to the SSD 
+running out of free pages to write to, which would degrade performance tremendously. 
+\cite{Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}
 
@@
 
-SSD manufacturers have made numerous advantage on their end to help mitigate what they are bad at. There are several 
+SSD manufacturers have made numerous advances on their end to help mitigate what SSDs are bad at. There are several 
 places apt for improvement, such as the FTL, or simply the amount of hardware included in the device. The more efficient 
@@
 An improved interface for the device will also increase SSD performance. Currently, SSDs use the same interface that 
-HDDs have employed, serial advanced technology attachment generation 3 (SATA-3), which is limited to somewhere around
+HDDs have employed, the Serial Advanced Technology Attachment generation 3 (SATA-3), which is limited to somewhere around
 600 MB/sec of thourughput. While that may be sufficent for HDDs, SSDs can easily outpace that. \cite{Micheloni2013}
 
-Fortunately, there are faster interfaces present on modern day systems. The peripheral component interconnect express 
+Fortunately, there are faster interfaces present on modern day systems. The Peripheral Component Interconnect Express 
 (PCIe) interface is used in all modern-day systems to connect important devices together. In fact, the SATA controller 
@@
 
-Systems that are data-heavy will need to attempt to optimize their access patterns for SSDs rather than HDDs.
+Systems that are data-heavy will need to attempt to optimize their access patterns for SSDs rather than HDDs. Knowing that 
+SSDs are slow at small, random writes, systems will want to avoid that behavior. Previous assumptions based on the workings 
+of HDDs can be disregarded.
 
